A Bayesian approach to  uncertainty in word embedding bias estimation

### 2023-09-06 paper/CLsubmission/bayesBiasCL.tex

In Section \ref{sec:two} we describe the two key measures discussed in
this paper, \textsf{WEAT} and \textsf{MAC}, explaining how they are
calculated and how they are supposed to work. In Section
\ref{sec:challenges} we first argue in Subsection
\ref{subsec:interpretability}, that it is far from clear how results
given in terms of \textsf{WEAT} or \textsf{MAC} are to be interpreted.
Second, in Subsection \ref{subsec:problems} we explain the statistical
problems that arise when one uses pre-averaged data in such contexts, as
these measures do. In Section \ref{sec:bayesian} we explain the
alternative Bayesian approach that we propose. In Section
\ref{sec:results} we elaborate on the results that it leads to,
including a somewhat skeptical view of the efficiency of debiasing
methods, discussed in Subsection \ref{subsec:rethinking}. Finally, in
Section \ref{sec:related} we spend some time placing our results in the
ongoing discussions.\footnote{\textbf{Disclaimer:} throughout the paper
  we will be mentioning and using word lists and stereotypes we did not
  formulate, which does not mean we condone any judgment made therein or
  underlying a given word selection. For instance, the Gender dataset
  does not recognize non-binary categories, and yet we use it without
  claiming that such categories should be ignored.}

@@

%\newpage
\appendix

@@

>>>>

In Section \ref{sec:two} we describe the two key measures discussed in
this paper, \textsf{WEAT} and \textsf{MAC}, explaining how they are
calculated and how they are supposed to work. In Section
\ref{sec:challenges} we first argue in Subsection
\ref{subsec:interpretability}, that it is far from clear how results
given in terms of \textsf{WEAT} or \textsf{MAC} are to be interpreted.
Second, in Subsection \ref{subsec:problems} we explain the statistical
problems that arise when one uses pre-averaged data in such contexts, as
these measures do. In Section \ref{sec:bayesian} we explain the
alternative Bayesian approach that we propose. In Section
\ref{sec:results} we elaborate on the results that it leads to,
including a somewhat skeptical view of the efficiency of debiasing
methods, discussed in Subsection \ref{subsec:rethinking}. Finally, in
Section \ref{sec:related} we spend some time placing our results in the
ongoing discussions.\footnote{\textbf{Disclaimer:} throughout the paper
  we will be mentioning and using word lists and stereotypes we did not
  formulate, which does not mean we condone any judgment made therein or
  underlying a given word selection. For instance, the Gender dataset
  does not recognize non-binary categories, and yet we use it without
  claiming that such categories should be ignored.}$^{, }$ \footnote{A few more philosophical comments on the enterprise of reflecting on bias in language models can be found in Appendix \ref{appendix:philosophical}.}

@@

%\newpage

\appendix

@@

\hypertarget{philosophical}{%
	\subsection{A philosophical commentary}}

\label{appendix:philosophical}

### 2023-09-06 paper/paperDraft.Rmd

A common approach in natural language processing is to  represent words  by  vectors of real numbers---such representations are called \emph{embeddings}.  One way to construct an embedding---we will focus our attention on non-contextual language models^[One example of a contextualized representation is BERT. Another is GPT.]---is to use  a large corpus to train  a  neural network to assign vectors to words in a way that optimizes for co-occurrence prediction accuracy. Such vectors can then be compared in terms of their similarity---the usual measure is cosine similarity---and the results of such comparisons can be used in downstream tasks. Roughly speaking, cosine similarity is an imperfect mathematical proxy for semantic similarity [@Mikolov2013efficient].

One response to the raising of the issue of bias in natural language models might be to say that there is not much point to reflecting on such biases, as they are  unavoidable. This unavoidability might seem in line with the arguments to the effect that learning algorithms are always value-laden [@JohnsonValueFree]: they employ inductive methods that require design-, data-, or risk-related decisions that have to be guided by extra-algorithmic considerations. Such choices necessarily involve value judgments and have to do, for instance, with what simplifications or risks one finds acceptable. 
Admittedly, algorithmic decision making cannot fulfill the value-free ideal, but this only means that even more attention needs to be paid to the values underlying different techniques and decisions, and to the values being pursued in a particular use of an algorithm.

@@

However, if our goal is to develop downstream tools that perform tasks that we care about without further perpetuating  or exacerbating harmful stereotypes, we still have good reasons to try to minimize the negative impact. Moreover, it is often not the case that the corpora mirror reality---to give a trivial example, heads are spoken of more often than kidneys, but this does not mean that kidneys occur much less often in reality than heads. To give a more relevant example, the disproportionate association of female words with female occupations in a corpus actually greatly exaggerates the actual lower disproportion in the real distribution of occupations [@gordon2012reporting].

>>>>

A common approach in natural language processing is to  represent words  by  vectors of real numbers---such representations are called \emph{embeddings}.  One way to construct an embedding---we will focus our attention on non-contextual language models^[One example of a contextualized representation is BERT. Another is GPT.]---is to use  a large corpus to train  a  neural network to assign vectors to words in a way that optimizes for co-occurrence prediction accuracy. Such vectors can then be compared in terms of their similarity---the usual measure is cosine similarity---and the results of such comparisons can be used in downstream tasks. Roughly speaking, cosine similarity is an imperfect mathematical proxy for semantic similarity [@Mikolov2013efficient].^[One response to the raising of the issue of bias in natural language models might be to say that there is not much point to reflecting on such biases, as they are  unavoidable. This unavoidability might seem in line with the arguments to the effect that learning algorithms are always value-laden [@JohnsonValueFree]: they employ inductive methods that require design-, data-, or risk-related decisions that have to be guided by extra-algorithmic considerations. Such choices necessarily involve value judgments and have to do, for instance, with what simplifications or risks one finds acceptable. 
Admittedly, algorithmic decision making cannot fulfill the value-free ideal, but this only means that even more attention needs to be paid to the values underlying different techniques and decisions, and to the values being pursued in a particular use of an algorithm.

@@

However, if our goal is to develop downstream tools that perform tasks that we care about without further perpetuating  or exacerbating harmful stereotypes, we still have good reasons to try to minimize the negative impact. Moreover, it is often not the case that the corpora mirror reality---to give a trivial example, heads are spoken of more often than kidneys, but this does not mean that kidneys occur much less often in reality than heads. To give a more relevant example, the disproportionate association of female words with female occupations in a corpus actually greatly exaggerates the actual lower disproportion in the real distribution of occupations [@gordon2012reporting].]
